---
title: "phonActivR: Complete Tutorial"
author: "Yu-Cheng Lin and Hitomi Kambara"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phonActivR: Complete Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
set.seed(1234)  # Ensures reproducible synthetic data across sessions
```

## Installation

`phonActivR` requires R (>= 4.0.0) and is installed from GitHub using the
`devtools` package. In your R console or RStudio terminal, run:

```{r install, eval=FALSE}
# Install devtools if not already present
install.packages("devtools")

# Install phonActivR from GitHub. build_vignettes = TRUE makes this
# tutorial available afterwards via vignette("phonActivR-tutorial")
devtools::install_github("yclinpsy/phonActivR", build_vignettes = TRUE)

# Load for the current session
library(phonActivR)
```

The package will automatically install its dependencies (`ggplot2`, `dplyr`,
`patchwork`, `cli`, `rlang`). No Java, no Python, no external configuration is
required.

## Overview

This vignette demonstrates the complete `phonActivR` workflow, from stimulus
specification through simulation, visualization, and empirical overlay, using
the built-in Japanese–English bilingual stimulus set as a worked example.

## Step 1: Load Package and Example Stimuli

```{r load}
library(phonActivR)

# Load built-in 42-item stimulus set
stim <- example_stimuli_jp()
stim
```

## Step 2: Run Simulation

```{r simulate}
sim <- run_simulation(stim, delta_values = c(0, 10, 20, 30))
summary(sim)
```

## Step 3: Visualize Predictions

All plotting functions share the package's publication theme
(`theme_phonactivr()`) and colorblind-safe Okabe–Ito palette
(`phonactivr_colors()`), and draw **no titles or subtitles inside the
image** — figure titles belong in your manuscript's captions, so every
figure below is journal-ready as produced. Pass a `title` argument only
if you want one (e.g., for slides), and use the theme on your own plots
with `+ theme_phonactivr()`.

### Competition Effect Curves

```{r fig1, fig.width=14, fig.height=6.5}
plot_competition(sim)
```

### Asymmetry Gradient

```{r fig2, fig.width=12, fig.height=6.5}
plot_asymmetry(sim)
```

### Onset Timing

```{r fig3, fig.width=11, fig.height=6.5}
plot_onset_timing(sim, mean_trial_ms = 1500)
```

## Step 4: Overlay Empirical Data

`example_empirical()` is the **recommended starting point** — it generates
realistic synthetic curves that run identically to real data, so you can
execute every line without external dependencies. Once you understand the
workflow, substitute your own data.frame (same columns, no other changes).

Note that `true_delta` encodes a specific competing hypothesis:
`true_delta = 0` is the universalist prediction; `true_delta > 0` is the
language-specific grain-size hypothesis (Cutler & Otake, 1994). Specifying this before data collection
constitutes quantitative pre-registration of the expected temporal effect
structure (cf. Henninger et al., 2025; BRM TOP Guidelines Level 2).

### Mouse-Tracking Example (Primary Demo Path)

```{r overlay_mt, fig.width=12, fig.height=6.5}
emp_mt <- example_empirical(sim, true_delta = 15,
  group_labels = c("JP bilinguals", "EN monolinguals"),
  paradigm = "mouse-tracking")

overlay_empirical(sim, emp_mt,
  group_colors = c("JP bilinguals" = "black",
                   "EN monolinguals" = "grey50"))
```

### Eye-Tracking Example (Primary Demo Path — second paradigm)

```{r overlay_et, fig.width=12, fig.height=6.5}
emp_et <- example_empirical(sim, true_delta = 12,
  group_labels = c("L2 learners", "L1 controls"),
  paradigm = "eye-tracking", noise_sd = 0.02, seed = 99)

overlay_empirical(sim, emp_et,
  group_colors = c("L2 learners" = "black",
                   "L1 controls" = "grey50"))
```

## Step 4b: Formal Model Selection (Goodness of Fit)

Visual overlay alone leaves the choice of the best-fitting delta to the eye.
`goodness_of_fit()` replaces this with a formal RSS/AIC model comparison,
run separately for each group in the empirical data. (When `fit = TRUE`,
the default, `overlay_empirical()` already runs this internally, reports
the selected delta for each group in the console, and attaches the full
AIC table to the returned plot as `attr(plot, "fit")` — nothing is drawn
inside the image itself.)

```{r gof}
gof <- goodness_of_fit(sim, emp_mt)
attr(gof, "best_delta")
```

Use the `time_window` argument to check that the selected delta is robust
to the analysis window (e.g., excluding the first 20% of the trial):

```{r gof_window}
gof_win <- goodness_of_fit(sim, emp_mt, time_window = c(21, 101))
attr(gof_win, "best_delta")
```

## Analysis Windows and Cross-Paradigm Comparability

The onset of measurable competition depends on the paradigm (eye-tracking
analyses conventionally begin ~200 ms after word onset; mouse-tracking
includes motor planning and execution time; ERP components are scored in
component-specific windows). Because phonActivR works on *normalized* time
(percent of the analysis window the researcher chose), the best-fitting
delta is defined relative to that window. Three practical rules follow:

1. **Within a paradigm**, best-fitting delta values are directly comparable
   across groups when trial durations and analysis windows are comparable.
2. **Across paradigms**, do not compare raw delta values. Instead compare
   each group's delta *against its own delta = 0 baseline* — that is,
   whether a prosodic constraint is present and how strong it is relative
   to the family of curves for that paradigm.
3. **Always report** the analysis window, and check robustness with
   `goodness_of_fit(..., time_window = ...)`.

## Step 5: Export Results

```{r export, eval=FALSE}
export_results(sim, prefix = "moraic_sim", dir = "output/")
```

## Using Your Own Data (The Researcher's End Goal)

Replace the synthetic data with your actual GAMM smooth estimates. The
bundled `example_gamm` dataset has **exactly** the structure of a typical
GAMM export (time, CV_effect, C_effect, CV_se, C_se, group), so the code
below is fully runnable as-is; to use your own data, replace
`data(example_gamm)` with `gamm_data <- read.csv("your_smooths.csv")` and
keep everything else unchanged.

```{r own_data, fig.width=12, fig.height=6.5}
# Bundled GAMM-format dataset (JP group generated at delta = 15,
# EN group at delta = 0 — see ?example_gamm)
data(example_gamm)
head(example_gamm)

# Convert the two component effects into the overlay format
emp_real <- data.frame(
  time      = example_gamm$time,
  asymmetry = example_gamm$CV_effect - example_gamm$C_effect,
  se        = sqrt(example_gamm$CV_se^2 + example_gamm$C_se^2),
  group     = example_gamm$group
)

overlay_empirical(sim, emp_real,
  group_colors = c("JP bilinguals" = "black",
                   "EN monolinguals" = "grey50"))

# Formal selection: JP group -> delta = 10 (nearest grid value to the
# generating delta = 15); EN group -> delta = 0 (its generating value)
goodness_of_fit(sim, emp_real)
```

## From Raw Data to the Overlay Format (GAMM users)

`overlay_empirical()` needs only a four-column data.frame: `time`
(normalized bins), `asymmetry` (the empirical CV--C difference), `se`
(optional), and `group` (optional). If you analyze your time-course data
with GAMMs, the difference smooth maps onto this directly (schematic;
adapt object names to your pipeline):

```{r gamm_stub, eval=FALSE}
# After fitting a GAMM to your mouse-tracking / eye-tracking data
# (e.g., with mgcv::bam), extract the CV-vs-C difference smooth per
# group with your usual tooling (e.g., itsadug::get_difference), then:
emp <- data.frame(
  time      = diff_smooth$time_bin,   # 1..101 normalized bins
  asymmetry = diff_smooth$difference, # estimated CV - C effect
  se        = diff_smooth$se,         # its standard error
  group     = diff_smooth$group      # e.g., "JP bilinguals"
)
overlay_empirical(sim, emp)
```

The bundled generation script `data-raw/make_example_gamm.R` documents
the exact format contract with runnable code.

A platform note: when saving figures with base `pdf()` on systems
without UTF-8 locales, the delta glyph in legends can trigger font
warnings; use `ggsave(..., device = cairo_pdf)` or PNG output.

## Cross-Linguistic Stimuli: Custom Phoneme Similarities

The built-in feature matrix covers 24 English consonants and 15 vowels.
Phonemes outside this set would silently receive a flat default similarity
(0.3) with every other phoneme — `run_simulation()` now warns when this
would happen. For such stimuli, directly enter pairwise similarity values
(0–1) with `custom_similarity()`; no full feature matrix is needed:

```{r custom_sim}
# Example: Korean tense stops, not in the built-in English set
my_pairs <- data.frame(
  p1         = c("k*", "k*", "t*"),
  p2         = c("k",  "g",  "t"),
  similarity = c(0.85, 0.60, 0.85)
)
sim_mat <- custom_similarity(my_pairs)
phoneme_similarity("k*", "k", similarity_matrix = sim_mat)

# Pass the same matrix to the full pipeline:
# run_simulation(stim, similarity_matrix = sim_mat)
```

## Parameter Sensitivity

```{r sensitivity, fig.width=12, fig.height=5}
gammas <- c(0.02, 0.04, 0.06)
sims <- lapply(gammas, function(g) {
  run_simulation(stim, gamma = g, verbose = FALSE)
})

# Compare: asymmetry gradient shape is robust to gamma
library(patchwork)
plots <- lapply(seq_along(gammas), function(i) {
  plot_asymmetry(sims[[i]],
    title = paste0("gamma = ", gammas[i]))
})
plots[[1]] + plots[[2]] + plots[[3]]
```

## Power Guidance and Study Planning

Before collecting data, use `power_guidance()` to assess whether your design
is capable of discriminating between competing delta hypotheses. The
`max_asymmetry` column in `summary(sim)` provides the expected peak effect
size at each delta value — use these to inform power calculations.

```{r power, eval=FALSE}
# How separable are delta = 0 (universalist) and delta = 30 (strong constraint)?
power_guidance(sim, n_items = 42, n_participants = 24)
```

**Interpretation (four-level failure framework; Magnuson, Mirman, & Harris, 2012; You & Magnuson, 2018):**

A failure to find the expected delta difference in your empirical data can
arise at four levels:

1. **Theory level** — the language-specific grain-size hypothesis is wrong
2. **Implementation level** — phonActivR's equations misrepresent the theory
3. **Parameter level** — the default alpha/gamma/decay do not apply to your population
4. **Linking-hypothesis level** — your paradigm does not map cleanly onto the competition-effect quantity

Run the sensitivity analysis (see Parameter Sensitivity section above) to probe
levels 3 and 4 before interpreting a null result as evidence against level 1.

## Pre-Registration Workflow

Generating delta prediction curves *before* data collection is a form of
**quantitative pre-registration** of effect structure. Unlike a conventional
directional pre-registration ("we predict a CV advantage"), this specifies
the *temporal shape* of the expected competition asymmetry at a particular
delta value. Steps:

1. Run `run_simulation()` on your planned stimulus set
2. Use `power_guidance()` to confirm your design can discriminate the target delta values
3. Register the `sim` object (or the exported CSV from `export_results()`) on OSF before data collection
4. After data collection, overlay with `overlay_empirical()` and report the best-fitting delta value
