---
title: "phonActivR: Complete Tutorial"
author: "[FILL IN: Your Name]"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phonActivR: Complete Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
set.seed(1234)  # Ensures reproducible synthetic data across sessions
```

## Installation

`phonActivR` requires R (>= 4.0.0) and is installed from GitHub using the
`devtools` package. In your R console or RStudio terminal, run:

```{r install, eval=FALSE}
# Install devtools if not already present
install.packages("devtools")

# Install phonActivR from GitHub
devtools::install_github("YOURUSERNAME/phonActivR")

# Load for the current session
library(phonActivR)
```

The package will automatically install its dependencies (`ggplot2`, `dplyr`,
`patchwork`, `cli`, `rlang`). No Java, no Python, no external configuration is
required.

## Overview

This vignette demonstrates the complete `phonActivR` workflow, from stimulus
specification through simulation, visualization, and empirical overlay, using
the built-in Japanese–English bilingual stimulus set as a worked example.

## Step 1: Load Package and Example Stimuli

```{r load}
library(phonActivR)

# Load built-in 42-item stimulus set
stim <- example_stimuli_jp()
stim
```

## Step 2: Run Simulation

```{r simulate}
sim <- run_simulation(stim, delta_values = c(0, 10, 20, 30))
summary(sim)
```

## Step 3: Visualize Predictions

### Competition Effect Curves

```{r fig1, fig.width=14, fig.height=6.5}
plot_competition(sim)
```

### Asymmetry Gradient

```{r fig2, fig.width=12, fig.height=6.5}
plot_asymmetry(sim)
```

### Onset Timing

```{r fig3, fig.width=11, fig.height=6.5}
plot_onset_timing(sim, mean_trial_ms = 1500)
```

## Step 4: Overlay Empirical Data

`example_empirical()` is the **recommended starting point** — it generates
realistic synthetic curves that run identically to real data, so you can
execute every line without external dependencies. Once you understand the
workflow, substitute your own data.frame (same columns, no other changes).

Note that `true_delta` encodes a specific competing hypothesis:
`true_delta = 0` is the universalist prediction; `true_delta > 0` is the
Language-Specific Grain Size hypothesis. Specifying this before data collection
constitutes quantitative pre-registration of the expected temporal effect
structure (cf. Henninger et al., 2025; BRM TOP Guidelines Level 2).

### Mouse-Tracking Example (Primary Demo Path)

```{r overlay_mt, fig.width=12, fig.height=6.5}
emp_mt <- example_empirical(sim, true_delta = 15,
  group_labels = c("JP bilinguals", "EN monolinguals"),
  paradigm = "mouse-tracking")

overlay_empirical(sim, emp_mt,
  group_colors = c("JP bilinguals" = "black",
                   "EN monolinguals" = "grey50"),
  title = "Mouse-Tracking: Predicted and Empirical CV-C Asymmetry")
```

### Eye-Tracking Example (Primary Demo Path — second paradigm)

```{r overlay_et, fig.width=12, fig.height=6.5}
emp_et <- example_empirical(sim, true_delta = 12,
  group_labels = c("L2 learners", "L1 controls"),
  paradigm = "eye-tracking", noise_sd = 0.02, seed = 99)

overlay_empirical(sim, emp_et,
  group_colors = c("L2 learners" = "black",
                   "L1 controls" = "grey50"),
  title = "Eye-Tracking: Predicted and Empirical CV-C Asymmetry")
```

## Step 5: Export Results

```{r export, eval=FALSE}
export_results(sim, prefix = "moraic_sim", dir = "output/")
```

## Using Your Own Data (The Researcher's End Goal)

Replace the synthetic data with your actual GAMM smooth estimates.
The call to `overlay_empirical()` is **identical** — only the `emp` object changes:

```{r own_data, eval=FALSE}
# From your GAMM analysis
gamm_data <- read.csv("Combined_smooth_data.csv")

jp <- gamm_data[gamm_data$group == "JP", ]
en <- gamm_data[gamm_data$group == "EN", ]

emp_real <- data.frame(
  time      = rep(1:101, 2),
  asymmetry = c(jp$CV_effect - jp$C_effect,
                en$CV_effect - en$C_effect),
  se        = c(sqrt(jp$CV_se^2 + jp$C_se^2),
                sqrt(en$CV_se^2 + en$C_se^2)),
  group     = rep(c("JP bilinguals", "EN monolinguals"), each = 101)
)

overlay_empirical(sim, emp_real)
```

## Parameter Sensitivity

```{r sensitivity, fig.width=12, fig.height=5}
gammas <- c(0.02, 0.04, 0.06)
sims <- lapply(gammas, function(g) {
  run_simulation(stim, gamma = g, verbose = FALSE)
})

# Compare: asymmetry gradient shape is robust to gamma
library(patchwork)
plots <- lapply(seq_along(gammas), function(i) {
  plot_asymmetry(sims[[i]],
    title = paste0("gamma = ", gammas[i]))
})
plots[[1]] + plots[[2]] + plots[[3]]
```

## Power Guidance and Study Planning

Before collecting data, use `power_guidance()` to assess whether your design
is capable of discriminating between competing delta hypotheses. The
`max_asymmetry` column in `summary(sim)` provides the expected peak effect
size at each delta value — use these to inform power calculations.

```{r power, eval=FALSE}
# How separable are delta = 0 (universalist) and delta = 30 (strong constraint)?
power_guidance(sim, n_items = 42, n_participants = 24)
```

**Interpretation (four-level failure framework, You & Magnuson 2018):**

A failure to find the expected delta difference in your empirical data can
arise at four levels:

1. **Theory level** — the Language-Specific Grain Size hypothesis is wrong
2. **Implementation level** — phonActivR's equations misrepresent the theory
3. **Parameter level** — the default alpha/gamma/decay do not apply to your population
4. **Linking-hypothesis level** — your paradigm does not map cleanly onto the competition-effect quantity

Run the sensitivity analysis (see Parameter Sensitivity section above) to probe
levels 3 and 4 before interpreting a null result as evidence against level 1.

## Pre-Registration Workflow

Generating delta prediction curves *before* data collection is a form of
**quantitative pre-registration** of effect structure. Unlike a conventional
directional pre-registration ("we predict a CV advantage"), this specifies
the *temporal shape* of the expected competition asymmetry at a particular
delta value. Steps:

1. Run `run_simulation()` on your planned stimulus set
2. Use `power_guidance()` to confirm your design can discriminate the target delta values
3. Register the `sim` object (or the exported CSV from `export_results()`) on OSF before data collection
4. After data collection, overlay with `overlay_empirical()` and report the best-fitting delta value
